Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks

### 2022-03-29 rmd/manuscript.Rmd
edit: manuscript

n_rep <- max(sim_main_result$n_rep)
n_scenario <- sim_main_result %>% 
  filter(sd_disturb_source > sd_disturb_lon) %>% 
  summarize(n = n_distinct(param_set)) %>% 
  pull(n)
n_sim <- sim_main_result %>% 
  mutate(n_sim = n_timestep + n_burnin + n_warmup) %>% 
  pull() %>% 
  unique()
```

@@

>>>>

n_rep <- max(sim_main_result$n_rep)

n_scenario <- sim_main_result %>% 
  filter(sd_disturb_source > sd_disturb_lon) %>% 
  summarize(n = n_distinct(param_set)) %>% 
  pull(n)

n_sim <- sim_main_result %>% 
  mutate(n_sim = n_timestep + n_burnin + n_warmup) %>% 
  pull() %>% 
  unique()
```

@@

The origin of variation in food chain length (FCL), the number of feeding links from the basal species to top predators, has intrigued ecologists for decades due to its importance in regulating top-down effects [@paceTrophicCascadesRevealed1999], energy flow [@wangBiodiversityEcosystemFunctioning2018], and contaminant concentrations in top predators that humans often consume [@kiddEffectsTrophicPosition1998]. Hypotheses regarding controls of food chain length can be organized into three broad categories. First, the resource availability hypothesis states that basal productivity controls food chain length because imperfect energy conversion through predator-prey interactions restricts energy available to higher trophic levels [@oksanenExploitationEcosystemsGradients1981]. Second, the disturbance hypothesis (or the dynamical stability hypothesis) predicts longer food chains are vulnerable to environmental perturbations (i.e., the dynamical constraint). Thus, habitats with frequent and/or severe disturbance should support fewer trophic levels [@pimmNumberTrophicLevels1977]. Lastly, the ecosystem size hypothesis posits that larger ecosystems should harbor longer food chains [@takimotoEffectsProductivityDisturbance2012; @teruiSpatialDisturbanceSynchrony2019; @postEcosystemSizeDetermines2000].

Among the three hypotheses, the ecosystem size hypothesis has been consistently supported in empirical explorations, likely because multiple mechanisms may underlie the positive association [@takimotoEnvironmentalDeterminantsFoodchain2013]. For example, large ecosystems have increased species richness (species-area relationship [@lomolinoEcologyMostGeneral2000]), buffer against disturbances [@saboRoleDischargeVariation2010], and have greater numbers of immigrant sources in the ecosystem with stable metapopulations [@takimotoEnvironmentalDeterminantsFoodchain2013]. Meanwhile, the theoretical basis for the other two hypotheses rest on implicit assumptions of linear food chains, which are often violated in natural systems. As such, the effects of disturbance and resource availability on food chain length have been found to be highly variable among ecosystem types and geographic regions [@takimotoEffectsProductivityDisturbance2012; @diehlIntraguildPreySuffer2001; @takimotoEnvironmentalDeterminantsFoodchain2013; @postLongShortFoodchain2002].

More recently, food web ecologists have begun to recognize that FCL rarely responds to a single driver, but instead is determined by interactions of multiple factors. For example, @wardMechanisticTheoryAquatic2017 showed that FCL was more responsive to increasing productivity in large ecosystems compared with small ecosystems, whereas FCL was more sensitive to size in low productivity systems. Such context dependent responses to extrinsic factors seem to be tightly coupled with the arrangement of trophic links within a food web. For example, omnivory (feeding at multiple trophic levels) is a strong intrinsic driver of food chain length [@takimotoEffectsProductivityDisturbance2012; @postProximateStructuralMechanisms2007; @woottonOmnivoryStabilityFreshwater2017]. When omnivory is absent in three-species communities, the trophic structure assumes a simple linear food chain with maximum length equal to 3 (**Figure 1D**). However, as omnivory increases, FCL decreases proportionally as the top predator eats lower in the food web [@wardMechanisticTheoryAquatic2017; @woottonOmnivoryStabilityFreshwater2017]. Theoretical work predicts that food webs are destabilized when omnivory is prevalent, leading to the loss of species and shorter food chains even in productive, large ecosystems. Conversely, weak omnivory stabilize the food web and increase persistence of species within a community, particularly at intermediate levels of productivity and high disturbance levels [@woottonOmnivoryStabilityFreshwater2017; @woottonManyWeakInteractions2016]. Due to the context-dependency of FCL to multiple drivers, and the potential for interactive effects, a unified framework of multiple food chain drivers is needed to better understand when and where FCL will respond to these extrinsic and intrinsic factors.

While recent research efforts of food web ecologists have substantially advanced our understanding of food chain drivers, most of the previous work has been conducted in simple, two-dimensional systems, such as lakes [@postEcosystemSizeDetermines2000; @doiResourceAvailabilityEcosystem2009] and oceanic islands [@takimotoEcosystemSizeNot2008; @pringlePredatorinducedCollapseNiche2019]. However, many natural habitats have a high degree of spatial complexity that cannot be represented simply by ecosystem size. Branching ecosystems, for example, are ubiquitous yet overlooked landscape structures (e.g., rivers, trees, caves, mountain ranges) [@grantLivingBranchesPopulation2007; @altermattDiversityRiverineMetacommunities2013]. While we know that food web interactions are spatially influenced [@calcagnoConstraintsFoodChain2011; @pillaiMetacommunityTheoryExplains2011; @takimotoEffectsProductivityDisturbance2012; @teruiSpatialDisturbanceSynchrony2019], the theory to account for it has lagged, particularly in branching habitat architectures. Here, we theoretically explore drivers of food chain length in rivers, an excellent example of branching ecosystems.

Ample evidence suggests that branching structure controls environmental heterogeneity and dispersal pathways of organisms within river ecosystems [@altermattDiversityRiverineMetacommunities2013; @tonkinRoleDispersalRiver2018]. Water and materials propagate downstream as small streams hierarchically join to form larger streams and rivers. Consequently, environmental signals are highly variable among tributaries while showing a strong within-tributary correlation e.g., flood disturbance [@nakamuraDisturbanceRegimesStream2000; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. Confluences serve as merging points where tributary dynamics aggregate to form distinct environmental conditions in downstream habitat patches. Habitat heterogeneity therefore increases with branching complexity (**Figure 1A**), which we define here as the probability of branching per unit river distance. Hence, highly-branched river networks provide diverse habitats that buffer the impact of large-scale environmental fluctuations, broadening opportunities for species to recolonize habitat patches post-disturbance [@teruiMetapopulationStabilityBranching2018]. Thus, ecosystem complexity in river networks may be a widespread yet overlooked extrinsic environmental driver which needs to be incorporated to understand food chain length.

We develop a model in which trophic community dynamics within branching habitats are simulated. We investigate extrinsic controls such as ecosystem size (number of patches), complexity (branching probability), and disturbance regime. In addition, our simulation model incorporates intrinsic drivers of food chain length including food web structure, productivity level, and dispersal ability, allowing us to investigate the context dependency on drivers of food chain length across extrinsic factors. We find that ecosystem size and complexity are both positively related to FCL, but the form of this relationship is dependent on other extrinsic and intrinsic factors. Our model is an important step in unifying drivers of FCL across extrinsic and intrinsic drivers and parsing their context-dependency in spatially complex ecosystems.
